Is Crime Really On A Rise In Toronto? An Analysis of The Past And Current Crime Rates of Toronto

### 2024-01-24 outputs/paper/paper.qmd
updated paper and scripts

---
title: "Is Crime Really On A Rise In Toronto? An Analysis of The Past And Current Crime Rates of Toronto"
author: 
  - Adrian Ly
thanks: "Code and data analysis are available at: https://github.com/AdrianUofT/Toronto-Crime-Rates-Investigation.git"
date: today
date-format: long
abstract: "Current mainstream news outlets are reporting a significant increase in crime rates, a trend that from our data analysis supports. Our findings indicate a dramatic increase in crimes such as homicide and auto theft, with them increasing over 210% from 2014. Compared to 2022, homicide rates have escalated by over 180% with auto theft having an increase of 20%. These substantial increases in crime rate underscore an evergrowing concern around public safety in Toronto and highlight the urgent need for effective crime prevention strategies."
format: pdf
number-sections: true
bibliography: references.bib
toc: true
---

@@

Over the recent years, major news outlets have been increasingly focused on the trends and patterns of crime rates in our city, Toronto. The narrative of a major increase in overall crime is sparking debate and concern over the general public's safety. This has caused a comprehensive analysis to understand the nature of these trends. To dissect and interpret crime data, this paper focuses on various types of crime to provide an objective view of how crime patterns and trends have evolved. More specifically, our analysis focuses on more petty thefts such as bike theft to more serious crimes such as homicide and auto theft. We examine the data from 2014 to 2023 seeking to uncover the underlying trends and percentage increases or decreases in crime rates. The goal is to offer a clear, data-driven perspective beyond anecdotal evidence and media narratives. Understanding these trends is not only important to inform the general public about them but also to point the direction for law enforcement and policymakers to make imperative choices to combat these criminals. This paper is structured into the following sections: Data, Results, Discussion, and Conclusion. In the data section I discuss the nature of these crime rates obtained from the City of Toronto's `opendatatoronto` [@citeOpenDataToronto] library, and the steps I took to clean and analyze the data. The result section highlights the trends found in the analysis process. The discussion section provides further evaluation of the trends and serves to provide insight as to the potential reasoning behind these crime trends and possible areas that require more attention. Lastly, the conclusion section is to summarize the main findings of this paper.

@@

# Results

@@

# Creating the line graph for average crime rates
ggplot(average_crime_rates_per_year, aes(x = year, y = average_rate, color = crime)) +
  geom_line() +
  labs(title = "Average Crime Rates by Type Across Years (2014-2023)",
       x = "Year",
       y = "Average Crime Rate",
       color = "Crime Type") +
  theme_minimal()
```

@@

@@

# Discussion

@@

# Conclusion

>>>>

---
title: "Is Crime Really On A Rise In Toronto? An Analysis of The Past And Current Crime Rates of Toronto"
author: 
  - Adrian Ly
thanks: "Code and data analysis are available at: https://github.com/AdrianUofT/Toronto-Crime-Rates-Investigation.git"
date: today
date-format: long
abstract: "Current mainstream news outlets are reporting a significant increase in crime rates, a trend that our data analysis supports. Our findings indicate a dramatic increase in crimes such as homicide and auto theft, with them increasing over 210% from 2014. Compared to 2022, homicide rates have escalated by over 180% with auto theft having an increase of 20%. These substantial increases in crime rate underscore an evergrowing concern around public safety in Toronto and highlight the urgent need for effective crime prevention strategies."
format: pdf
number-sections: true
bibliography: references.bib
toc: true
---

@@

Over the recent years, major news outlets have been increasingly focused on the trends and patterns of crime rates in our city, Toronto. The narrative of a major increase in overall crime is sparking debate and concern over the general public's safety. This has caused a comprehensive analysis to understand the nature of these trends. To dissect and interpret crime data, this paper focuses on various types of crime to provide an objective view of how crime patterns and trends have evolved. More specifically, our analysis focuses on more petty thefts such as bike theft to more serious crimes such as homicide and auto theft. We examine the data from 2014 to 2023 seeking to uncover the underlying trends and percentage increases or decreases in crime rates. The goal is to offer a clear, data-driven perspective beyond anecdotal evidence and media narratives. Understanding these trends is not only important to inform the general public about them but also to point the direction for law enforcement and policymakers to make imperative choices to combat these criminals. This paper is structured into the following sections: Data, Results, Discussion, and Conclusion. In @sec-data, I discuss the nature of these crime rates obtained from the City of Toronto's `opendatatoronto` [@citeOpenDataToronto] library, and the steps I took to clean and analyze the data. @sec-results highlights the trends found in the analysis process. @sec-discussion provides further evaluation of the trends and serves to provide insight as to the potential reasoning behind these crime trends and possible areas that require more attention. Lastly, @sec-conclusion is to summarize the main findings of this paper.

@@

# Results {#sec-results}

@@

# Changing the names on the graph legend
average_crime_rates_per_year$crime <- factor(average_crime_rates_per_year$crime,
                                             levels = c("assault", "autotheft", "biketheft", "breakenter", "homicide", "robbery", "shooting", "theftfrommv", "theftover"),
                                             labels = c("Assault", "Auto Theft", "Bike Theft", "Break and Enter", "Homicide", "Robbery", "Shooting", "Theft from Motor Vehicle", "Theft over"))

# Creating the line graph with updated legend names
ggplot(average_crime_rates_per_year, aes(x = year, y = average_rate, color = crime)) +
  geom_line() +
  labs(title = "Average Crime Rates by Type Across Years (2014-2023)",
       x = "Year",
       y = "Average Crime Rate",
       color = "Crime Type") +
  theme_minimal() 
```

@@

# Changing the names on the graph legend
percentage_change_data$crime <- factor(percentage_change_data$crime,
                                             levels = c("assault", "autotheft", "biketheft", "breakenter", "homicide", "robbery", "shooting", "theftfrommv", "theftover"),
                                             labels = c("Assault", "Auto Theft", "Bike Theft", "Break and Enter", "Homicide", "Robbery", "Shooting", "Theft from Motor Vehicle", "Theft over"))

@@

# Discussion {#sec-discussion}

@@

# Conclusion {#sec-conclusion}
